---
title: "hw02"
author: "Aisha Uduman"
date: "September 19, 2017"
output: 
  html_document: 
    keep_md: yes
---

```{r}
library(gapminder)
library(tidyverse)
```

## Exploring the Gapminder object 
Justify each answer by calling the appropriate R functions 

#### Is it a data frame, matrix, vector or list? 

Gapminder is a data frame, as it contains a list of vectors of equal length, or same number of rows. 

```{r}
summary(gapminder)
```


#### What's its class?

```{r}
sapply(gapminder,class)
```

Classes are as follows:
* country = factor
* continent = factor 
* year = integer 
* lifeExp = numeric 
* pop = integer 
* gdpPercap = numeric


#### How many variables/columns?

```{r}
ncol(gapminder)
```

6 variables/columns 

#### How many rows/observations?

```{r}
nrow(gapminder)
```
1704 rows/observations

#### Can you get these facts about extent or size in more than one way? Can you imagine different functions being useful in different contexts? 


#### What data type is each variable? 

```{r}
head(gapminder)
```

```{r}
typeof("Afghanistan")
typeof("Asia")
typeof(1952)
typeof(28.801)
typeof(8425333)
typeof(779.4453)
```

The variables are as follows: 
* country = character
* continent = character 
* year = double
* lifeExp = double
* pop = double
* gdpPercap = double 

### Exploring individual variables 

