---
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(RColorBrewer))
```



# Factor Management 

The goals for this homework: 
- Define factor variables 
- Drop factor/levels 
- Reorder levels based on knowledge from data 



### Gapminder version: 

I decided to stick with the Gapminder dataset as I find it much easier to use intuitively (especially the column names) compared to the Singer database, and also because this part of the class is getting to be a bit over my head at times, so keeping it simple is key :ok_hand:



**Drop Oceania** 

I'm first filtering by continent, to exclude Oceania. 

```{r}
gap_noocean <- gapminder %>% 
  filter(continent != "Oceania")
```

Comparing this to the original Gapminder dataframe, which has 5 continents and 1704 rows, 24 rows being associated with Oceania. 
Checking to see if Oceania has been dropped: We can see that now Oceania has 0 values, but *still shows up in the summary for this new dataframe.* 

```{r}
summary(gapminder)
nrow(gapminder)
levels(gapminder$continent)

summary(gap_noocean)
```

We need to remove Oceania from this dataframe altogether. To do so, I used the `droplevels()` argument, to drop the unused levels: 

```{r}
gap_noocean_drop <- gap_noocean %>%
  droplevels()

nlevels(gap_noocean_drop$continent)
summary(gap_noocean_drop)
```

Now we can see that Oceania no longer exists within the continent variable, and the number of levels within this variable is 4 (as opposed to 5 in the original gapminder). Since Oceania was associated with 24 rows in the original Gapminder, we should now see this new dataframe having 24 less rows: 

```{r}
nrow(gap_noocean_drop)
```

1704 - 1680 = 24. Yay! 


Just to confirm that only 4 continents are present: 

```{r}
levels(gap_noocean_drop$continent)
```

Great, Oceania no longer exists in this new dataframe. 



**Reorder the levels of `country` or `continent`** 

Reordering countries based on ascending life expectancies: 

```{r}
fct_reorder(gap_noocean_drop$country, gap_noocean_drop$lifeExp) %>% 
  levels()
```


Reordering countries based on descending minimum GDP per capita: 
```{r}
fct_reorder(gap_noocean_drop$country, gap_noocean_drop$gdpPercap, min, .desc = TRUE) %>% 
  levels()
```


Reordering continents based on descending variance in GDP per capita
```{r}
fct_reorder(gap_noocean_drop$continent, gap_noocean_drop$gdpPercap, var, .desc = TRUE) %>% 
  levels()
```

For a more focussed example, let's look at the variable life expectancy, focusing on the countries with the lowest life expectancies. 

```{r}
country_reorder_lifeExp <- fct_reorder(gap_noocean_drop$country, gap_noocean_drop$lifeExp, min, .desc = FALSE) %>% 
  levels() %>% 
  head()

head(country_reorder_lifeExp)
```

We can see that now that the reordering has been done, Rwanda, Afghanistan, Gambia, Angola, Sierra Leone and Cambodia have the lowest life expectancies in the gapminder dataframe. 


Now I want to define the life expectancies as a dataframe with only the country, year and its corresponding life expectancy to see if it's changing over time: This will get rid of the population, continent, and GDP per capita factors. 

```{r}
min_lifeExp <- gap_noocean_drop %>% 
  filter(country %in% country_reorder_lifeExp) %>% 
  select(year, country, lifeExp)
```


We can plot this to visualize the differences and trends over time: 

```{r}
plot_minlifeExp <- ggplot(min_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
plot_minlifeExp
```

Cool, now we can see the output of our reordered factors selecting for minimum life expectancy. 


# Common Part 

Characterize the derived data before and after your factor re-leveling. 
Exploring effects of reordering a factor and factor reordering coupled with `arrange()`.


To see how our countries are organized (most likely alphebetically): 


```{r}
min_lifeExp %>% 
  group_by(country)

kable(min_lifeExp, digits=2)
```



Yes, the countries are arranged alphabetically. I want them to be displayed in terms of minimum life expectancy, so the dataframe must be re-arranged to display life expectancy levels in ascending order. *This is a note to myself, if I wanted to arrange by descending order I would use `arrange(desc(lifeExp))`, as `arrange()` automatically uses ascending order. 


```{r}
ascending_lifeExp <- min_lifeExp %>% 
  group_by(country) %>% 
  arrange(lifeExp)
kable(ascending_lifeExp, digits =2)
```


This is what I wanted, now we can see our dataframe is sorted in terms of increasing life expectancy, and not country or year. I want to see if this has any effect on a figure: 


```{r}
ggplot(ascending_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
```

No, unless I'm doing it wrong...the visualizations are exactly the same! Even the countries in the legend are still in alphabetical order. I played around with this for a long time and found that nothing I did with `arrange()` made a change in the resulting plot. 



# Visualization design 




# File I/O 

Writing and reading to file 

**Writing:** 
```{r}
write.csv(ascending_lifeExp, file = "ascending_lifeExp.csv")
```

I can see that the .csv file shows up in the homework 5 folder. 

**Reading:** 

Now to read the .csv file I just wrote, I will use the `read_csv` function: 

```{r}
read_csv_asc <- read_csv("ascending_lifeExp.csv")
kable(read_csv_asc, digits=2)
```

It works, I can write a .csv file and pull it up again. 




# Writing Figures to File 

Using `ggsave`, I will save a plot made earlier in this homework: 

```{r}
ggsave("MinLifeExpPlot.pdf", plot_minlifeExp, width = 8, height = 6)
```

I played around a bit with figure height and width. After running, I noticed the .pdf immediately appearing in my homework 5 folder, found ![here](https://github.com/auduman/stat545-hw-uduman-aisha/blob/master/hw05/MinLifeExpPlot.pdf)


# Reporting my Progress 

- I spent too long trying to make sure I only selected a few of the countries with lowest expectancy, and realised instead of using `head()` to view them and then manually select them like I've been doing before, I can just pipe `head()` after the `levels()` argument to only retain the first 6 levels that I want to visualize. 

- I had issues figuring out how to enter my pathway to save my csv files....using the `file=""` I realised it saves automatically to the folder that this markdown file is in (so my homework 5 folder), but what if I wanted to save it somewhere else? I kept getting error messages, so need to look into that later. 


**Useful Links**

- Helpful information on [`arrange()`](https://www.r-bloggers.com/dplyr-example-1/)
