---
output: 
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(RColorBrewer))
```



# Factor Management 

The goals for this homework: 

- Define factor variables 
- Drop factor/levels 
- Reorder levels based on knowledge from data 



### Gapminder version: 

I decided to stick with the Gapminder dataset as I find it much easier to use intuitively (especially the column names) compared to the Singer database, and also because this part of the class is getting to be a bit over my head at times, so keeping it simple is key :ok_hand:



**Drop Oceania** 

First let's see the summary for Gapminder: 

```{r}
summary(gapminder)
nrow(gapminder)
levels(gapminder$continent)
```

We can see that Gapminder originally has 5 continents and 1704 rows, 24 rows being associated with Oceania. 
Now I'm filtering by continent, to exclude Oceania. 

```{r}
gap_noocean <- gapminder %>% 
  filter(continent != "Oceania")
```


Checking to see if Oceania has been dropped: We can see that now Oceania has 0 values, but *still shows up in the summary for this new dataframe.* 

```{r}
summary(gap_noocean)
```


We need to remove Oceania from this dataframe altogether. To do so, I used the `droplevels()` argument, to drop the unused levels: 

```{r}
gap_noocean_drop <- gap_noocean %>%
  droplevels()

nlevels(gap_noocean_drop$continent)
summary(gap_noocean_drop)
```

Now we can see that Oceania no longer exists within the continent variable, and the number of levels within this variable is 4 (as opposed to 5 in the original gapminder). Since Oceania was associated with 24 rows in the original Gapminder, we should now see this new dataframe having 24 less rows: 

```{r}
nrow(gap_noocean_drop)
```

1704 - 1680 = 24. Yay! 



Just to confirm that only 4 continents are present: 

```{r}
levels(gap_noocean_drop$continent)
```


Great, Oceania no longer exists in this new dataframe! 



**Reorder the levels of `country` or `continent`** 

Reordering countries based on ascending life expectancies: 

```{r}
fct_reorder(gap_noocean_drop$country, gap_noocean_drop$lifeExp) %>% 
  levels()
```


Reordering continents based on descending variance in GDP per capita:

```{r}
fct_reorder(gap_noocean_drop$continent, gap_noocean_drop$gdpPercap, var, .desc = TRUE) %>% 
  levels()
```


For a more focussed example, let's look at the variable life expectancy, focusing on the countries with the lowest life expectancies. I used `head`() to identify the 6 countries with lowest life expectancies.

```{r}
country_reorder_lifeExp <- fct_reorder(gap_noocean_drop$country, gap_noocean_drop$lifeExp, min, .desc = FALSE) %>% 
  levels() %>% 
  head()

head(country_reorder_lifeExp)
```

We can see that now that the reordering has been done, Rwanda, Afghanistan, Gambia, Angola, Sierra Leone and Cambodia have the lowest life expectancies in the gapminder dataframe. 


Now I want to define the life expectancies as a dataframe with only the country, year and its corresponding life expectancy to see if it's changing over time: This will get rid of the population, continent, and GDP per capita factors. 

```{r}
min_lifeExp <- gap_noocean_drop %>% 
  filter(country %in% country_reorder_lifeExp) %>% 
  select(year, country, lifeExp)
```


We can plot this to visualize the differences and trends over time: 

```{r}
plot_minlifeExp <- ggplot(min_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
plot_minlifeExp
```

Cool, now we can see the output of our reordered factors selecting for minimum life expectancy. 




# Common Part 


Characterize the derived data before and after your factor re-leveling. 


**Exploring effects of reordering a factor and factor reordering coupled with `arrange()`.**


To see how our countries are organized (most likely alphebetically): 


```{r}
min_lifeExp %>% 
  group_by(country)
```

Yes, the countries are arranged alphabetically. I want them to be displayed in terms of minimum life expectancy, so the dataframe must be re-arranged to display life expectancy levels in ascending order. *This is a note to myself, if I wanted to arrange by descending order I would use `arrange(desc(lifeExp))`, as `arrange()` automatically uses ascending order. 


```{r}
ascending_lifeExp <- min_lifeExp %>% 
  group_by(country) %>% 
  arrange(lifeExp)
```


To view only the first 15 rows of this also very long dataset, I used the `head()` function: 

```{r}
head(ascending_lifeExp, 15) %>% 
kable()
```


This is what I wanted, now we can see our dataframe is sorted in terms of increasing life expectancy, and not country or year. Does arranging the data have any effect on a figure? 

```{r}
ggplot(ascending_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
```

It doesn't seem like this has any effect on a graph. 




**Exploring the effects of reordering a factor and factor reordering coupled with `arrange()`**

I want to see if coupling the factor reordering with `arrange()` has any effect on a figure: 

With `arrange()`: 

```{r}
ggplot(ascending_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
```

Without `arrange(): 

```{r}
plot_minlifeExp <- ggplot(min_lifeExp, aes(year, lifeExp, colour=country)) + 
  facet_wrap(~country) + 
  geom_line() + 
  scale_color_brewer(palette="Accent") + 
  labs(x = "Year", y = "Life Expectancy", title = "Countries with Minimum Life Expectancies") + theme_bw()
plot_minlifeExp
```


No, unless I'm doing it wrong...the visualizations are exactly the same! Even the countries in the legend are still in alphabetical order. I played around with this for a long time and found that nothing I did with `arrange()` made a change in the resulting plot. 





# Visualization design 

I think it will be interesting to take one of the first (slightly complex) plots I ever made in ggplot, which is probably hideous, and try and work on it to make it look more visually pleasing after learning a lot about visualization techniques in this class. This is an initial plot of 

```{r}
ggplot(gapminder, aes(year, lifeExp, colour = continent)) +
    facet_wrap(~ continent) +
    geom_point() + 
  geom_smooth(method='loess', colour = "black", se=FALSE, span = 5) 
```

Now to make it look nicer: I adjusted the figure height and width, changed the alpha transparency to 0.4, made the line dashed. I enjoyed manually selecting colours, however I know I'm not an expert so I decided to use one of the R Colour Brewer palettes for qualitative data....to be honest I really don't like any of them, but "Set2" seemed okay. 

In addition, I also used `labs` and `theme` to adjust the axis title, positions, font sizes, the strip background colour and text colour, removed the legend title as I think it's pretty self-explanatory. I also used the `scales::pretty_breaks` argument to adjust the y-axis as I found ticks representing 20 years of age not to be super helpful. 


```{r, fig.width=8, fig.height=6}
ggplot(gapminder, aes(year, lifeExp, colour = continent)) +
    facet_wrap(~ continent) +
    geom_point(alpha=0.4) + 
  scale_y_continuous(breaks = scales::pretty_breaks(n = 7)) + 
  geom_smooth(method='loess', colour = "black", se=FALSE, span = 5, linetype = "dashed") +
  labs(x = "Year", y = "Life Expectancy", title = "Life Expectancy of Continents over Time") + 
  scale_color_brewer(palette = "Set2") + 
  theme_bw() +
    theme(strip.background = element_rect(fill="royalblue4"), 
          strip.text = element_text(size=12, face="bold", colour = "white"), 
          plot.title = element_text(size=14,face="bold", hjust = 0.5),
          axis.title = element_text(size=12, face="bold"), 
          legend.title = element_blank())
```




# File I/O 


Writing and reading to file 


**Writing:** 

```{r}
write.csv(ascending_lifeExp, file = "ascending_lifeExp.csv")
```

I can see that the .csv file shows up in the homework 5 folder. 


**Reading:** 

Now to read the .csv file I just wrote, I will use the `read_csv` function. I also used the `head()` function again to only show the first 15 rows to save space. 

```{r}
read_csv_asc <- read_csv("ascending_lifeExp.csv")

head(read_csv_asc, 15) %>% 
kable(digits=2)
```

It works, I can write a .csv file and pull it up again. 




# Writing Figures to File 

Using `ggsave`, I will save a plot made earlier in this homework: 

```{r}
ggsave("MinLifeExpPlot.pdf", plot_minlifeExp, width = 8, height = 6)
```

I played around a bit with figure height and width. After running, I noticed the .pdf immediately appearing in my homework 5 folder, found ![here](https://github.com/auduman/stat545-hw-uduman-aisha/blob/master/hw05/MinLifeExpPlot.pdf)


I'll try saving a plot object made from homework 3: 

```{r}
p <- ggplot(gapminder %>% 
  group_by(continent,year) %>% 
  summarise(mean_lifeExp=mean(lifeExp)), 
  aes(year, mean_lifeExp, group=continent, colour=continent)) + 
  geom_smooth(method="loess", se=FALSE) + 
  scale_colour_manual(values=c("darkslateblue", "olivedrab", "tan3", "orangered3", "maroon4")) +
  labs(x = "Year", y = "Life Expectancy", title = "Life Expectancy Over Time for the Continents") + 
  theme(axis.title = element_text(size=14),
        axis.text.x = element_text(size=12),
        axis.text.y = element_text(size=12),
        plot.title = element_text(size=16, hjust = 0.5))
```

```{r}
ggsave("my_plot.pdf", p)
```

I can see that this created another plot in my homework 5 folder, which can be found ![here](https://github.com/auduman/stat545-hw-uduman-aisha/blob/master/hw05/object_p.pdf). To be honest, I can't think of when this actually matters compared to the previous `ggsave` function I ran before. I have to read up about this more. 


# Reporting my Progress 

I found this week's homework to take quite a bit of time - I was also rushed due to my TA commitments and a trip I had to take on the weekend. The actual work took quite a long time so I wasn't able to spend as much time as I would have liked on the aesthetics of the output. 

**Some issues I encountered:**

- I spent too long trying to make sure I only selected a few of the countries with lowest expectancy, and realised instead of using `head()` to view them and then manually select them like I've been doing before, I can just pipe `head()` after the `levels()` argument to only retain the first 6 levels that I want to visualize. 

- I had issues figuring out how to enter my pathway to save my csv files....using the `file=""` I realised it saves automatically to the folder that this markdown file is in (so my homework 5 folder), but what if I wanted to save it somewhere else? I kept getting error messages, so need to look into that later. 

- I would like to learn how to display two plots side by side (i.e. if I could show the plots of factor reordering and factor reordering + `arrange()`) it would look much better and also be easier to notice any differences. I noticed Vincenzo posted [this](http://stat545.com/block020_multiple-plots-on-a-page.html) link, however I ran out of time to try it. 


**Useful Links**

- Helpful information on [`arrange()`](https://www.r-bloggers.com/dplyr-example-1/)

- A useful forum on using `scales::pretty_breaks(n=)` to change the number of [axis ticks.](https://stackoverflow.com/questions/11335836/increase-number-of-axis-ticks) 

- [ggplot2](http://sape.inf.usi.ch/quick-reference/ggplot2/colour) quick reference for colour and fill.

- Notes from STAT 545 class on ['being the boss of your factors'](http://stat545.com/block029_factors.html) were super useful to keep wrapping my head around these concepts. 
